---
title: "Pivotting"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Pivotting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

This vignette describes the use of the new `pivot()` function. 

There are two primary ways that `pivot()` differs from previous functions in tidyr, reshape2, and reshape:

* The reshaping operation is specified with a data frame that describes 
  precisely how metadata stored in column names becomes real data variables. 
  This is inspired by the [cdata](https://winvector.github.io/cdata/) package 
  by John Mount and Nina Zumel.
  
* Reshaping works with multiple measure variables that may have different types. 
  This is inspired by the enhanced `melt()` and `dcast()` functions provided 
  by the [data.table](https://github.com/Rdatatable/data.table/wiki) package
  by Matt Dowle and Arun Srinivasan.

```{r setup}
library(tidyr)
library(dplyr)
library(readr)
```

# Wide to long

## Pew data

```{r}
pew <- read_csv("pew.csv", col_types = list())
pew
```

To tidy this data we first create a spec. We are tidying from wide to long so we use `pivot_spec_long()`:

```{r}
spec <- pew %>% pivot_spec_long(-1, "count", "income")
spec
```

* The first argument describes which columns this transformation applies to.
  In this case, it's every column apart from the first.

* The `cell_val` gives the name of the variable stored in the cells. 
  In this the case the values are measurements of counts.
  
* The `col_name` gives the name of the variable stored in the column names.
  In this case, it's an `income` range.

We can now achieve the desired reshaping by pivot `pew` with this `spec`:

```{r}
pew %>% pivot(spec)
```

## Billboard

Similar to `pew`, but the variable encoded in the columns is not a string, and requires some additional manipulation to extract:

```{r}
billboard <- read_csv("billboard.csv", col_types = list()) %>%
  mutate_at(vars(starts_with("wk")), as.numeric)
billboard
```

```{r}
billboard %>% pivot_spec_long(starts_with("wk"), "rank", "week")
```

```{r}
spec <- billboard %>%
  pivot_spec_long(starts_with("wk"), "rank", "week") %>%
  mutate(week = parse_number(week))
spec
```

```{r}
billboard %>% pivot(spec)
```

## WHO tuberculosis data

In more complex cases, the column name might encode multiple variables, and require significant parsing to extract.

```{r}
who
```

```{r}
who %>% pivot_spec_long(new_sp_m014:newrel_f65)

spec <- who %>%
  pivot_spec_long(new_sp_m014:newrel_f65, "count") %>%
  extract(variable, c("diagnosis", "gender", "age"), "new_?(.*)_(.)(.*)") %>%
  mutate(age = factor(age, levels = unique(age), ordered = TRUE))
spec
```

```{r}
who %>% pivot(spec, na.rm = TRUE)
```

## Multiple value columns 

From <https://cran.r-project.org/web/packages/data.table/vignettes/datatable-reshape.html>

```{r}
family <- tibble::tribble(
  ~family_id, ~age_mother,  ~dob_child1,  ~dob_child2, ~gender_child1, ~gender_child2,
          1L,         30L, "1998-11-26", "2000-01-29",             1L,             2L,
          2L,         27L, "1996-06-22",           NA,             2L,             NA,
          3L,         26L, "2002-07-11", "2004-04-05",             2L,             2L,
          4L,         32L, "2004-10-10", "2009-08-27",             1L,             1L,
          5L,         29L, "2000-12-05", "2005-02-28",             2L,             1L,
) %>%
  mutate_at(vars(starts_with("dob")), parse_date)

spec <- family %>%
  pivot_spec_long(-(1:2)) %>%
  separate(variable, c("measure", "child")) %>%
  mutate(child = parse_number(child))

family %>%
  select(-age_mother) %>%
  pivot(spec, na.rm = TRUE)
```

## Census data

Lightly modified from Table 5 "completions" found at 
<https://www.census.gov/construction/nrc/index.html>.

```{r}
construction <- read_csv("construction.csv", col_types = list("2 to 4 units" = col_integer()))
construction
# multiple variables encoded, but no way to parse; must manually specify

spec <- tribble(
  ~col_name,         ~units,  ~region,      ~measure,
  "Total",            "total", NA,          "n",
  "1 unit",           "1",     NA,          "n",
  "2 to 4 units",     "2-4",   NA,          "n",
  "5 units or more",  "5+",    NA,          "n",
  "Northeast",        NA,      "Northeast", "n",
  "Midwest",          NA,      "Midwest",   "n",
  "South",            NA,      "South",     "n",
  "West",             NA,      "West",      "n",
)
spec

construction %>% pivot(spec)
```

# Long to wide

# Multiple transformations

## Weather data

```{r}
# need to pivot in both directions

weather <- read_csv("weather.csv", col_types = list()) 

weather %>% pivot_spec_long(d1:d31, "temp", "day")

weather_spec <- weather %>%
  pivot_spec_long(d1:d31, "temp", "day") %>%
  mutate(day = parse_number(day))
weather_spec
weather2 <- pivot(weather, weather_spec, na.rm = TRUE)
weather2

spec2 <- weather2 %>% pivot_spec_wide(element, temp)
spec2
weather2 %>% pivot(spec2)
```

## Contact list

Requires to widening twice.

Inspired by <https://github.com/jienagu/tidyverse_examples/blob/master/example_long_wide.R>

```{r}
contacts <- 
  c(
    "company", "Toyota", "name", "Jiena McLellan",
    "company", "McDonalds", "name", "John Smith"
  )

df <- tibble(meta = contacts)
df <- df %>% 
  mutate(
    id = cumsum(meta == "company"),
    row = rep(seq_len(nrow(df) / 2), each = 2),
    type = rep(c("var", "value"), length = nrow(df))
  ) 

spec <- pivot_spec_wide(df, type, meta)
df2 <- df %>% pivot(spec)

spec2 <- pivot_spec_wide(df2, var, value)
df2 %>% select(-row) %>% pivot(spec2)

```

