---
title: "Pivotting"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Pivotting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

This vignette describes the use of the new `pivot()` function. 

There are two primary ways that `pivot()` differs from previous functions in tidyr, reshape2, and reshape:

* The reshaping operation is specified with a data frame that describes 
  precisely how metadata stored in column names becomes real data variables. 
  This is inspired by the [cdata](https://winvector.github.io/cdata/) package 
  by John Mount and Nina Zumel. For simple cases, this specification is
  implicit, but for more complex cases it is useful to make it explicit.
  
* Reshaping works with multiple measure variables that may have different types. 
  This is inspired by the enhanced `melt()` and `dcast()` functions provided 
  by the [data.table](https://github.com/Rdatatable/data.table/wiki) package
  by Matt Dowle and Arun Srinivasan.

```{r setup}
library(tidyr)
library(dplyr)
library(readr)
```

# Wide to long

## Pew data

```{r}
pew <- read_csv("pew.csv", col_types = list())
pew
```

To tidy this data we first create a spec. We are tidying from wide to long so we use `pivot_long_spec()`:

```{r}
spec <- pew %>% pivot_long_spec(-1, names_to = "count", values_to = "income")
spec
```

* The first argument describes which columns this transformation applies to.
  In this case, it's every column apart from the first.

* The `cell_val` gives the name of the variable stored in the cells. 
  In this the case the values are measurements of counts.
  
* The `col_name` gives the name of the variable stored in the column names.
  In this case, it's an `income` range.

We can now achieve the desired reshaping by pivot `pew` with this `spec`:

```{r}
pew %>% pivot_long(spec = spec)
```

## Billboard

Similar to `pew`, but the variable encoded in the columns is not a string, and requires some additional manipulation to extract:

```{r}
billboard <- read_csv("billboard.csv", col_types = list()) %>%
  mutate_at(vars(starts_with("wk")), as.numeric)
billboard
```

```{r}
billboard %>% pivot_long_spec(starts_with("wk"), names_to = "week", values_to = "rank")
```

```{r}
spec <- billboard %>%
  pivot_long_spec(starts_with("wk"), names_to = "week", values_to = "rank") %>%
  mutate(week = parse_number(week))
spec
```

```{r}
billboard %>% pivot_long(spec = spec)
```

## WHO tuberculosis data

In more complex cases, the column name might encode multiple variables, and require significant parsing to extract.

```{r}
who
```

```{r}
who %>% pivot_long_spec(new_sp_m014:newrel_f65)

spec <- who %>%
  pivot_long_spec(new_sp_m014:newrel_f65, values_to = "count") %>%
  extract(name, c("diagnosis", "gender", "age"), "new_?(.*)_(.)(.*)") %>%
  mutate(age = factor(age, levels = unique(age), ordered = TRUE))
spec
```

```{r}
who %>% pivot_long(spec = spec, na.rm = TRUE)
```

## Multiple value columns 

From <https://cran.r-project.org/web/packages/data.table/vignettes/datatable-reshape.html>

```{r}
family <- tibble::tribble(
  ~family_id, ~age_mother,  ~dob_child1,  ~dob_child2, ~gender_child1, ~gender_child2,
          1L,         30L, "1998-11-26", "2000-01-29",             1L,             2L,
          2L,         27L, "1996-06-22",           NA,             2L,             NA,
          3L,         26L, "2002-07-11", "2004-04-05",             2L,             2L,
          4L,         32L, "2004-10-10", "2009-08-27",             1L,             1L,
          5L,         29L, "2000-12-05", "2005-02-28",             2L,             1L,
) %>%
  mutate_at(vars(starts_with("dob")), parse_date)

spec <- family %>%
  pivot_long_spec(-(1:2)) %>%
  separate(name, c("measure", "child")) %>%
  mutate(child = parse_number(child))

family %>%
  select(-age_mother) %>%
  pivot_long(spec = spec, na.rm = TRUE)
```

## Census data

Lightly modified from Table 5 "completions" found at 
<https://www.census.gov/construction/nrc/index.html>.

```{r}
construction <- read_csv("construction.csv", col_types = list("2 to 4 units" = col_integer()))
construction
# multiple variables encoded, but no way to parse; must manually specify

spec <- tribble(
  ~col_name,         ~units,  ~region,      ~measure,
  "Total",            "total", NA,          "n",
  "1 unit",           "1",     NA,          "n",
  "2 to 4 units",     "2-4",   NA,          "n",
  "5 units or more",  "5+",    NA,          "n",
  "Northeast",        NA,      "Northeast", "n",
  "Midwest",          NA,      "Midwest",   "n",
  "South",            NA,      "South",     "n",
  "West",             NA,      "West",      "n",
)
spec

construction %>% pivot_long(spec = spec)
```

# Long to wide

## Single measure

From <http://stackoverflow.com/questions/24929954>

```{r}
df <- expand_grid(
    Product = c("A", "B"), 
    Country = c("AI", "EI"), 
    Year = 2000:2014
  ) %>%
  mutate(value = rnorm(nrow(.))) %>%
  filter((Product == "A" & Country == "AI") | (Product == "B" & Country == "EI"))

spec <- df %>% 
  expand(Product, Country, measure = "value") %>% 
  mutate(
    col_name = paste0(Product, "_", Country)
  )
spec

df %>% pivot_wide(spec = spec) %>% head()
```

## Multiple measure

(From Jenny Bryan)

```{r}
input <- tribble(
  ~hw,   ~name,  ~mark,   ~pr,
  "hw1", "anna",    95,  "ok",
  "hw1", "alan",    90, "meh",
  "hw1", "carl",    85,  "ok",
  "hw2", "alan",    70, "meh",
  "hw2", "carl",    80,  "ok"
)

spec <- input %>% 
  expand(hw, measure = c("mark", "pr")) %>% 
  mutate(
    col_name = paste0(hw, "_", measure)
  )
spec

input %>% pivot_wide(spec = spec)
```



# Multiple transformations

## Anscombe's quartet

Need to introduce identifying variable

```{r}
anscombe <- anscombe %>% 
  as_tibble() %>% 
  mutate(id = row_number())
anscombe

spec <- anscombe %>% 
  pivot_long_spec(-id) %>% 
  separate(name, c("var", "ex"), 1, convert = TRUE)
spec

anscombe %>% 
  pivot_long(spec = spec) %>% 
  pivot_wide(var, value)
```

## Weather data

```{r}
# need to pivot in both directions

weather <- read_csv("weather.csv", col_types = list()) 

weather %>% pivot_long_spec(d1:d31, "temp", "day")

weather_spec <- weather %>%
  pivot_long_spec(d1:d31, names_to = "day", values_to = "temp") %>%
  mutate(day = parse_number(day))
weather_spec
weather2 <- pivot_long(weather, spec = weather_spec, na.rm = TRUE)
weather2

spec2 <- weather2 %>% pivot_wide_spec(element, temp)
spec2
weather2 %>% pivot_wide(spec = spec2)
```

## Contact list

Requires to widening twice.

Inspired by <https://github.com/jienagu/tidyverse_examples/blob/master/example_long_wide.R>

```{r}
contacts <- 
  c(
    "company", "Toyota", "name", "Jiena McLellan",
    "company", "McDonalds", "name", "John Smith"
  )

df <- tibble(meta = contacts)
df <- df %>% 
  mutate(
    id = cumsum(meta == "company"),
    row = rep(seq_len(nrow(df) / 2), each = 2),
    type = rep(c("var", "value"), length = nrow(df))
  ) 

spec <- pivot_wide_spec(df, names_from = type, values_from = meta)
df2 <- df %>% pivot_wide(spec = spec)

spec2 <- pivot_wide_spec(df2, names_from = var, values_from = value)
df2 %>% select(-row) %>% pivot_wide(spec = spec2)
```

