---
title: "Programming with tidyr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Programming with tidyr}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

```{r setup, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(collapse = T, comment = "#>")
options(tibble.print_min = 4L, tibble.print_max = 4L)
set.seed(1014)

# Manually "import"; only needed for old dplyr which uses old tidyselect
# which doesn't attach automatically in tidy-select contexts
all_of <- tidyselect::all_of
```

## Introduction

Most tidyr verbs use **tidy evaluation** in some way. Tidy evaluation is a special type of non-standard evaluation used throughout the tidyverse. There are two basic forms found in tidyr:

* `drop_na()`, `fill()`, `pivot_longer()`/`pivot_wider()`, `nest()`/`unnest()`, 
  `separate()`/`extract()`, and `unite()`, use **tidy selection** so you can 
  easily choose variables based on their position, name, or type 
  (e.g. `1:3`, `starts_with("x")` or `is.numeric`).

* `expand()`, `crossing()` and `nesting()` use **data masking** so that you can 
  use data variables as if they were variables in the environment (i.e. you 
  write `my_variable` not `df$myvariable`). 
  
Since tidy selection is the most common, we'll focus on it in this vignette. You can learn more about data masking function in the equivalent vignette in dplyr: <https://dplyr.tidyverse.org/dev/articles/programming.html>. For other considerations when writing tidyr code in package, please see `vignette("in-packages")`.

Tidy selection and data masking make interactive data exploration fast and fluid, but they add some new challenges when you attempt to use them indirectly such as in a for loop or a function. This vignette shows you how to overcome those challenges. We'll first go over the basics of tidy selection and data masking, talk about how to use them indirectly, and then show you a number of recipes to solve common problems.

Before we go on, we'll attach the packages we use, expose the version of tidyr, and make a small dataset to use in examples.

```{r, message = FALSE}
library(tidyr)

packageVersion("tidyr")

mini_iris <- as_tibble(iris)[c(1, 2, 51, 52, 101, 102), ]
mini_iris
```

## Tidy selection

Underneath all functions that use tidy selection is the [tidyselect](http://tidyselect.r-lib.org/) package. It provides a miniature domain specific language that makes it easy to select columns by name, position, or type. For example:

* `select(df, 1)` selects the first column; 
  `select(1, last_col())` selects the last column.
  
* `select(df, c(a, b, c))` selects columns `a`, `b`, and `c`.

* `select(df, starts_with("a"))` selects all columns whose name starts with "a";
  `select(df, ends_with("z"))` selects all columns whose name ends with "z".
  
* `select(df, is.numeric)` selects all numeric columns.

You can see more details in `?tidyr_tidy_select`.

### Indirection

Tidy selection makes a common task easier at the cost of making a less common task harder. When you want to use tidy select indirectly with the column specification stored in an intermediate variable, you'll need to learn some new tools. There are three main cases where this comes up:

*   When you have the tidy-select specification in a function argument, you 
    must **embrace** the argument by surrounding it in doubled braces.

    ```{r}
    nest_egg <- function(df, cols) {
      nest(df, egg = {{ cols }})
    }
    
    nest_egg(mini_iris, -Species)
    ```

*   When you have a character vector of variable names, you must use `all_of()` 
    or `any_of()` depending on whether you want the function to error if a 
    variable is not found. These functions allow you to write for loops or a
    function that takes variable names as a character vector.
    
    ```{r}
    nest_egg <- function(df, cols) {
      nest(df, egg = all_of(cols))
    }
    
    vars <- c("Sepal.Length", "Sepal.Width", "Petal.Length", "Petal.Width")
    nest_egg(mini_iris, vars)
    ```

*   In more complicated cases, you might want to use tidyselect directly:

    ```{r}
    sel_vars <- function(df, cols) {
      tidyselect::eval_select(rlang::enquo(cols), df)
    }
    sel_vars(mini_iris, -Species)
    ```
    
    Learn more in `vignette("tidyselect")`.
    
Note that many tidyr functions use `...` so you can easily select many variables, e.g. `fill(df, x, y, z)`. I now believe that the disadvantages of this approach outweigh the benefits, and that this interface would have been better as `fill(df, c(x, y, z))`. For new functions that select columns, please just use a single argument and not `...`.
