---
title: "Rectangling"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{rectangling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

Rectangling is the art and craft of taking a deeply nested list (often sourced from wild caught JSON or XML) and taming it into a tidy data set of rows and columns. There are three functions from tidyr that are particularly useful for rectangling:

* `unnest_longer()` takes each element of a list-column and makes a new row.
* `unnest_wider()` takes each element of a list-column and makes a new column.
* `hoist()` is similar to `unnest_wider()` put only plucks out selected 
  components, and can reach down multiple levels.
  
A very large number of data rectangling problems can be solved by combining this functions with a splash of dplyr code. To illustrate these techniques, we'll use the repurrrsive package, which provides a number of datasets with deeply nested lists originally captured from JSON APIs.
  
```{r setup, message = FALSE}
library(tidyr)
library(dplyr)
library(repurrrsive)
```

## GitHub users

We'll start with `gh_users`, a list which contains information about six GitHub users. To begin, we put the `gh_users` list into a data frame:

```{r}
users <- tibble(user = gh_users)
```

This seems a bit counter-intuitive: why is the first step in making a list more to make it more complicated? But a data frame has a big advantage: it bundles together multiple vectors so that everything is tracked together in a single object.

Each `user` is a named list, which we want to turn into columns.

```{r}
names(users$user[[1]])
```

There are two ways to turn the list components into columns. `unnest_wider()` takes every component and makes a new column:

```{r}
users %>% unnest_wider(user)
```

But in this case, there are many components and we don't need most of them so we can instead use `hoist()`. `hoist()` allows us to pull out selected components using the same syntax as `purrr::pluck()`:

```{r}
users %>% hoist(user, 
  followers = "followers", 
  login = "login", 
  url = "html_url"
)
```

`hoist()` removes the named components from the `user` list-column, so you can think of it as moving components out of the inner list into the top-level data frame. 

## Github repos

We start off `gh_repos` similarly, by putting it in a tibble:

```{r}
repos <- tibble(user = gh_repos)
repos
```

In this case, the elements of `user` are a list of repositories that belong to that user. These should become new rows, so we use `unnest_longer()` rather than `unnest_wider()`:

```{r}
repos %>% unnest_longer(user, values_to = "repo")
```

In this case, the index of the repo within the user is not informative, so we can drop with `indices_to = NULL`:

```{r}
repos <- repos %>% unnest_longer(user, values_to = "repo", indices_to = NULL)
repos
```

Then we can use `unnest_wider()` or `hoist()`:

```{r}
repos %>% hoist(repo, 
  login = c("owner", "login"), 
  name = "name",
  homepage = "homepage",
  watchers = "watchers_count"
)
```

Note the use of `c("owner", "login")`: this allows us to reach two levels deep inside of a list. An alternative approach would be to pull out just `owner` and then put each element of it in a column:

```{r}
repos %>% 
  hoist(repo, owner = "owner") %>% 
  unnest_wider(owner)
```


## Game of Thrones characters

`got_chars` has a similar structure to `gh_users`: it's a list of named lists, which each element of the inner list describing some attribute of a GoT character. We start in the same way, first by creating a data frame and then by unnesting each component into a column:

```{r}
chars <- tibble(char = got_chars)
chars

chars2 <- chars %>% unnest_wider(char)
chars2
```

This is more complex than `gh_users` because some component of `char` are themselves a list, giving us a collection of list-columns:

```{r}
chars2 %>% select_if(is.list)
```

What you do next will depend on the purposes of the analysis. Maybe you want a row for every book and tv series that the character appears in:

```{r}
chars2 %>% 
  select(name, books, tvSeries) %>% 
  pivot_longer(c(books, tvSeries), names_to = "media", values_to = "value") %>% 
  unnest_longer(value, indices_to = NULL)
```

Or maybe you want to build a table that lets you match title to name:

```{r}
chars2 %>% 
  select(name, titles) %>% 
  unnest_longer(titles, values_to = "title", indices_to = NULL)
```

(Note that the empty titles (`""`) are due to an infelicity in the input `got_chars`: ideally people without titles would have a title vector of length 0, not a title vector of length 1 containing an empty string.)

## Sharla Gelfand's discography

We'll finish off with the most complex list [Sharla Gelfand's](https://sharla.party/posts/discog-purrr/). We'll start the usual way: putting the list into a single column data frame, and then widening so each component is a column. I also parse the `date_added` column into a real date-time (I'd normally use `readr::parse_datetime()`, but I can't here because it's a vignette and I don't want to introduce another dependency to tidyr).

```{r}
discs <- tibble(disc = discog) %>% 
  unnest_wider(disc) %>% 
  mutate(date_added = as.POSIXct(strptime(date_added, "%Y-%m-%dT%H:%M:%S"))) 
discs
```

At this level, we see information about when each disc was added to Sharla's discography, not any information about the disc itself. To do that we need to widen the `basic_information` column:

```{r, error = TRUE}
discs %>% unnest_wider(basic_information)
```

Unfortunately that fails because there's an `id` column inside `basic_information`. We can quickly see what's going on by setting `names_repair = "unique"`:

```{r}
discs %>% unnest_wider(basic_information, names_repair = "unique")
```

The problem is that `basic_information` repeats the `id` column that's also stored at the top-level, so we can just drop that:

```{r}
discs %>% 
  select(-id) %>% 
  unnest_wider(basic_information)
```

Alternatively, we could use `hoist()`:

```{r}
discs %>% 
  hoist(basic_information,
    title = "title",
    year = "year",
    label = list("labels", 1, "name"),
    artist = list("artists", 1, "name")
  )
```

Here I quickly extract the name of the first label and artist by indexing deeply into the nested list. You might want to do this more systematically using `unnest_wider()` and `unnest_longer()`

```{r}
discs %>% 
  hoist(basic_information, label = "labels") %>% 
  select(id, label) %>% 
  unnest_longer(label, values_to = "label", indices_to = NULL) %>% 
  unnest_wider(label, names_repair = "unique")
```

